# Task answers {#sec-tasks-answers}

::: callout-note
This page contains answers to the *Task* activities presented in selected chapters of the book.
:::


## From Chapter 2

### Task 1 {.unnumbered}

The completed data description table for the Citibike stations dataset.

<div style="font-size:0.8em;">

| Variable name  | Variable value   | Measurement level |
|----------------|------------------|-------------------|
| `name`         | "Central Park"   |  Categorical-nominal     |
| `capacity`     |  80              |         Ratio          |
| `rank_capacity`|  45              |      Ordinal     |
| `date_opened`  |  "2014-05-23"    |        Interval           |
| `longitude`    | -74.00149746     |         Interval          |
| `latitude`     | 40.74177603      |           Interval        |

</div>

### Task 2 {.unnumbered}


```{r}
#| echo: true
#| eval: false
# # A tibble: 7 × 3
#   day   Customer Subscriber
#   <ord>    <dbl>      <dbl>
# 1 Sun     0.198       0.144
# 2 Mon     0.137       0.163
# 3 Tue     0.144       0.172
# 4 Wed     0.104       0.125
# 5 Thu     0.0973      0.122
# 6 Fri     0.135       0.138
# 7 Sat     0.185       0.136
```

Code for exploring the proportion of trips made by day of week on `user_type`:

```{r}
#| echo: true
#| eval: false
ny_temporal |> 
  group_by(day, user_type) |> 
  summarise(count=sum(count)) |> ungroup() |> 
  pivot_wider(names_from=user_type, values_from=count) |> 
  mutate(across(.cols=c(Customer, Subscriber), .fns=~.x/sum(.x)))
```


## From Chapter 3


### Task 1 {.unnumbered}

The completed encoding description table for the Washington Post election map (@fig-wp-swing).

<div style="font-size:0.8em;">

| Data item         | Measurement level | Visual mark  | Visual channel | Rank         |
|-------------------|-------------------|--------------|----------------|--------------|
| County location | Interval   | Lines | Position in x- y-   |  1 mag:order |
| County winner             | Categorical-nominal             | Lines       | Colour hue          |  2 id:category    |
| County flip from 2012           | Categorical-ordinal             | Lines       | 1D size          | 3 mag:order       |
| County Swing fom 2012            | Ratio             | Lines        | Tilt/angle          | 4 mag:order        |
| State 'winner'             |   Categorical-nominal           | Polygon       | Colour hue          | 2 id:category       |


</div>

### Task 2 {.unnumbered}

Code to produce a set of histograms similar to those in @fig-hist-region.

```{r}
#| echo: true
#| eval: false
data_gb |>
  ggplot(mapping=aes(swing_con_lab)) +
  geom_histogram(fill="#003c8f") +
  # Annotate with median Swing.
  geom_vline(xintercept=4.44, size=.3)+
  labs(x="Swing", y="count")+
  facet_wrap(~region)
```


### Task 3 {.unnumbered}

The code to reproduce the graphic in @fig-scatters-con.

```{r}
#| echo: true
#| eval: false

con <- "#0575c9"
lab <- "#ed1e0e"
other <- "#bdbdbd"

data_gb |>
  mutate(is_flipped=seat_change_1719=="Conservative gain from Labour",
         is_flipped=if_else(is.na(is_flipped), FALSE, is_flipped),
         winner_19=case_when(
           winner_19 == "Conservative" ~ "Conservative",
           winner_19 == "Labour" ~ "Labour",
           TRUE ~ "Other"
         )) |>
  ggplot(aes(x=con_17, y=con_19)) +
  geom_point(aes(colour=winner_19, alpha=is_flipped, shape=is_flipped)) +
  geom_abline(intercept = 0, slope = 1, size=.3) +
  scale_colour_manual(values=c(con,lab,other)) +
  scale_alpha_ordinal(range=c(.5,1)) +
  scale_shape_manual(values=c(21,19)) +
  scale_x_continuous(limits=c(0,90)) +
  labs(x="vote share 2017 ", y="vote share 2019")
```


## From Chapter 4

### Task 1 {.unnumbered}

The code to reproduce the heatmap in the left column of @fig-borough-freqs-resids.


```{r}
#| echo: true
#| eval: false
# Vector of vehicles in order they appear in graphic. 
# Convert vehicle_type to factor to effect ordering.
order_type <- c("Car", "Taxi", "Bus", "Motorcycle", "Other","Van", "HGV", "Bicycle") 
# For new is_inner variable used to facet plot.
inner_boroughs <- c("Camden", "Greenwich", "Hackney", "Hammersmith and Fulham", "Islington",
                    "Kensington and Chelsea", "Lambeth", "Lewisham", "Southwark", "Tower Hamlets",
                    "Wandsworth", "Westminster", "City of London")
# Staged dataset where signed-chi residuals are created.
model_data <- ped_veh |> 
  filter(police_force == "Metropolitan Police" | police_force == "City of London") |> 
  mutate(
    is_inner=
      if_else(local_authority_district %in% inner_boroughs, "inner", "outer"),
    vehicle_type=factor(vehicle_type, levels=order_type)
  ) |> 
  group_by(local_authority_district) |> 
  mutate(row_total=n()) |> ungroup() |> 
  group_by(vehicle_type) |> 
  mutate(col_total=n()) |> ungroup() |> 
  mutate(grand_total=n()) |> 
  group_by(local_authority_district, vehicle_type) |> 
  summarise(
    observed=n(), row_total=first(row_total),  col_total=first(col_total), 
    grand_total=first(grand_total), expected=(row_total*col_total)/grand_total, 
    resid=(observed-expected)/sqrt(expected),
    is_inner=first(is_inner)
  ) |>  ungroup() 
# Find maximum residual value to ensure colour scheme is symmetrical on 0.
max_resid <- max(abs(model_data$resid))

# Plot heatmap.
model_data |>   
  ggplot(aes(x=vehicle_type, y=reorder(local_authority_district, row_total))) +
  geom_tile(aes(fill=resid), colour="#ffffff", size=.4) +
  facet_grid(is_inner~., scales="free_y", space="free_y") +
  scale_fill_distiller(palette="RdBu", direction=-1, limits=c(-max_resid,max_resid)) +
  guides(fill="none")
```


### Task 2 {.unnumbered}

The code to reproduce the design challenge plots in @fig-age-imd-light.

```{r}
#| echo: true
#| eval: false
# Calculate crash freqs by IMD class of location, dark/daylight and casulty age.
plot_data <- ped_veh |>
  filter(
    age_of_casualty>0, crash_quintile != "Data missing or out of range",  
    light_conditions != "Data missing or out of range"
    ) |>
  mutate(is_daylight=
           factor(if_else(light_conditions == "Daylight", "daylight", "dark"),
                  levels=c("dark", "daylight"))) |>
  group_by(age_of_casualty, is_daylight, crash_quintile) |>
  summarise(count=n()) |> ungroup() 

# Top plot.
plot_data |>
  ggplot(aes(x=age_of_casualty, y=count)) +
  geom_col(aes(colour=is_daylight), width=1) +
  facet_grid(is_daylight~crash_quintile, space="free_y", scales="free_y", 
             labeller=labeller(c("daylight", "dark"))) +
  scale_colour_manual(values=c("#08519c", "#c6dbef"), guide="none")+
  labs(y="crash count in hundreds", x="casualty age") +
  scale_y_continuous(
    breaks=c(c(2,4,6,8,10)*100),
    labels = scales::comma_format(scale = .01))

# Bottom plot.
plot_data |> 
  # Calcuate % crashes in daylight and expected daylight counts 
  # in each data item from this.
  mutate(total=sum(count)) |> 
  pivot_wider(names_from=is_daylight, values_from=count) |> 
  mutate(
    prop_daylight=sum(daylight, na.rm=TRUE)/first(total),
    expected_daylight=(daylight+dark)*prop_daylight
  ) |> 
  pivot_longer(cols=c(dark, daylight), names_to="is_daylight", values_to="count") |> 
  # Plot.
  ggplot(aes(x=age_of_casualty, y=count)) +
  geom_col(aes(colour=is_daylight), width=1) +
  geom_line(aes(y=expected_daylight, group=crash_quintile), colour="#737373", linewidth=.4) +
  facet_wrap(~crash_quintile, labeller=labeller(c("daylight", "dark")), nrow=1) +
  scale_colour_manual(values=c("#08519c", "#c6dbef"), guide="none")+
  labs(y="crash count in hundreds", x="casualty age") +
  scale_y_continuous(
    breaks=c(c(4,8,12)*100), 
    labels = scales::comma_format(scale = .01))
```


## From Chapter 5

### Task 1 {.unnumbered}

* For jobs filled in the City of London (CoL) from which borough does the largest number of workers commute? 
  + Answer: Wandsworth (Wnd)

* For jobs filled in Camden (Cmd) from which borough does the largest number of workers commute? 
  + Answer: Barnet (Barnt), maybe Islington (Isl)

* Eyeballing the graphic, identify the top 3 boroughs which appear to have the most localised labour markets in terms of in-commuting? 
  + Answer: Bexleyheath (Bxl), Havering (Hvr), Barking and Dagenham (BaD). Also, Sutton (Sttn), Grenwich (Grn) look very localised.


## From Chapter 6

The code to reproduce the annotated parallel-coordinate-plot in @fig-pcps.

```{r}
#| echo: true
#| eval: false
# List of variables ordered ascending by correlation with leave.
order_vars <- cons_data |>
  mutate(across(c(younger:heavy_industry), ~(.x-mean(.x))/sd(.x))) |> 
  pivot_longer(cols=younger:heavy_industry, names_to="expl_var", values_to="prop") |> 
  group_by(expl_var) |>  
  summarise(cor=cor(leave,prop)) |> ungroup() |> arrange(cor) |>  
  pull(expl_var)

# Staged dataset of z-score transformed variables and variables identifying 
# extreme Leave/Remain constituencies for highlighting.
plot_data <- cons_data |> 
  mutate(
    majority=if_else(leave>.5, "Leave", "Remain"),
    across(c(leave, younger:heavy_industry), ~(.x-mean(.x))/sd(.x)),
    decile=ntile(leave, 10), is_extreme=decile > 9 | decile < 2
  )  |> 
  select(
    majority, is_extreme, decile, constituency_name, leave, degree, professional, 
    younger, eu_born, no_car, white, own_home, christian, not_good_health, heavy_industry
  ) |>  
  # Change polarity in selected variables.
  mutate(degree=-degree, professional=-professional, younger=-younger, 
    eu_born=-eu_born, no_car=-no_car) |>  
  pivot_longer(cols= c(leave:not_good_health), names_to="var", values_to="z_score") |> 
  # Explanatory variable as factor ordered according to known assocs with Leave.
  mutate(
    var=factor(var, levels=c("leave", order_vars)),
    var=fct_rev(var)
  ) 
# Sample extreme constituencies -- for Leave and Remain -- each time plot is built.
annotate_data <- plot_data |> 
  filter(is_extreme) |> 
  group_by(decile) |> 
  sample_n(1) |> pull(constituency_name)

plot_data |>  
  ggplot(aes(x=var, y=z_score, colour=majority, group=c(constituency_name))) +
  geom_path(alpha=0.15, linewidth=.2) +
  # Highlight extreme remain/leave constituencies.
  geom_path(
    data= . %>% filter(constituency_name %in% annotate_data),
    alpha=1, linewidth=.4
  )+
  geom_text(
    data= . %>% filter(constituency_name %in% annotate_data, var=="leave"),
    aes(x="leave", y=z_score, label=str_wrap(constituency_name,15)), 
    size=3.5, vjust="top", hjust="centre", nudge_x=+.5
    ) +
  scale_colour_manual(values=c("#b2182b","#2166ac")) +
  coord_flip()
```