---
title: "Chapter 01 Template"
author: "Roger Beecham"
format: html
---

## Introduction

For each chapter of *vis4sds* there is an accompanying `.qmd` template, which provides scaffolding for completing the technical elements of each chapter and the accompanying additional *Tasks*.

## Setup

The code block below points to *the* core package you will use throughout the book. You may have already installed the `tidyverse`, with `install.packages("tidyverse")` and so simply need to make it available in session with `library(<package-name>)`.

```{r}
#| echo: true
#| eval: false
# Bundle of packages for doing modern data analysis
library(tidyverse)
```

## Plotting Anscombe's Quartet

There was reasonably little in the way of data analysis in this first chapter. However, to demonstrate how code and analysis can be executed in `.qmd` files, we can plot the famous [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet) datasets. When pointing to the `tidyverse` package, you actually automatically loaded these data. Should you wish to inspect them, type `View(anscombe)` in the R Console.

Below is some code for generating scatterplots of the Anscombe datasets. No need to worry about the code for the time being, but it consists of a bit of data reshaping using [`tidyr`](https://tidyr.tidyverse.org/), combined with a [`ggplot2`](https://ggplot2.tidyverse.org/) specification for generating the scatterplots.

```{r}
#| echo: true
#| eval: false
anscombe |> 
  pivot_longer(
    everything(),
    names_to = c(".value", "dataset"), names_pattern = "(.)(.)"
    ) |>
  ggplot(aes(x, y))+
  geom_point(colour="#003c8f", fill="#003c8f", pch=21) +
  stat_smooth(method=lm, se=FALSE, size=0.6, colour="#636363")+
  facet_wrap(~dataset, nrow=2)
```

If you're feeling particularly motivated, and want to extend the Anscombe example, you may wish to play around with [`datasauRus`](https://itsalocke.com/datasaurus/) package.

When you work with Quarto documents it is cumbersome to constantly render the document. Most often you will want to run each code chunk separately so that the results can be explored in your Environment. To do this, simply click on the green arrow to the top right each code block whenever you wish to run a section of code.

## Writing plot to your project folder

If you've followed the instructions in the chapter, you should have created an RStudio project and saved this `.qmd` file in you project's `reports` folder. Let's now save our Anscombe quartet plot to the project's `figures` folder by **evaluating** the code block below when you next click *Render* (HINT: you will need to edit `eval: <true|false>`).

```{r}
#| echo: true
#| eval: false
# Save plot
ggsave("./figures/anscombe.png", plot=plot,width=7, height=4, dpi=300)
```
