---
title: "Chapter 03 Template"
author: "Roger Beecham"
format: html
---

## Introduction

Reopen your project and save this `.qmd` file in the top-level folder. When working through the materials from this chapter run each code chunk separately by clicking the small green arrow beside each code chunk. By default I have set `eval: false` on most code chunks, so when you knit the code chunk will not execute.

The technical element involves analysing data from the 2019 UK General Election. You will work with results data recorded by Parliamentary Constituency. After importing and describing the dataset, you will generate data graphics that expose patterns in voting behaviour by writing ggplot2 specifications.

## Setup

The code block below points to the required packages. If you have been following the chapters sequentially, the only new package to install is [`parlitools`](https://cran.r-project.org/web/packages/parlitools/index.html).

```{r}
#| echo: true
#| eval: false
library(tidyverse) 
library(sf) 
library(here) 

# Uncomment to install
# install.packages("parlitools") 
library(parlitools)
```

## Import

Installing `parlitools` brings down the 2019 UK General Election dataset, along with other constituency-level datasets. Loading it with `library(parlitools)` makes these data available to your R session. The dataset containing 2019 UK General Election data is called `bes_2019`. This contains results data released by [House of Commons Library](https://commonslibrary.parliament.uk/research-briefings/cbp-8749/).

We also need to import geographic boundary data for Parliamentary Constituencies. This can be collected from [ONS Open Geography Portal](https://geoportal.statistics.gov.uk/). I have generated a simplified version of these geometries using [mapshaper](https://github.com/mbloch/mapshaper) and posted to the data repository. Run the code below to import into your R session.

```{r}
#| echo: true
#| eval: false
url <- "https://vis4sds.github.io/data/ch3/constituency_boundaries.geojson"
download.file(url, here("data", "constituency_boundaries.geojson"))
cons_outline <- st_read(here("data", "constituency_boundaries.geojson"), crs=27700)
```

To generate a version of the 2019 data limited to GB, and with the Swing variable (`swing_con_lab`) run the code below.

```{r}
#| echo: true
#| eval: false
data_gb <- bes_2019 |>
  filter(region != "Northern Ireland") |>
  mutate(
    # Recode to 0 Chorley incoming speaker, Buckingham outgoing speaker.
    swing_con_lab=if_else(constituency_name %in% c("Chorley", "Buckingham"), 0, 0.5*((con_19-con_17)-(lab_19-lab_17))
      )
  )
```

## Plot Con-Lab Swing graphic

Towards the end of the technical section, code for generating a map of Conservative-Labour Swing is described. Although the plot grammar is reasonably straightforward, the plot depends on a couple of helper functions (below). These can be loaded into your environment by executing the code below.

```{r}
#| echo: true
#| eval: false
url <- "https://vis4sds.github.io/data/ch3/spoke_helpers.R"
source(url)
```

The ggplot2 specification:

```{r}
#| echo: true
#| eval: false
# Find the maximum Swing values, to pin the min and max angles to.
max_shift <- max(abs(data_gb |> pull(swing_con_lab)))
min_shift <- -max_shift

# Define party_colours to contain just three values: hex codes for
# Conservative, Labour and Other.
# Define colours.
con <- "#0575c9"
lab <- "#ed1e0e"
other <- "#bdbdbd"
party_colours <- c(con, lab, other)
names(party_colours) <- c("Conservative", "Labour", "Other")

# Plot Swing map.
data_gb |>
  mutate(
    is_flipped=seat_change_1719 %in%
       c("Conservative gain from Labour","Labour gain from Conservative"),
    elected=if_else(
      !winner_19 %in% c("Conservative", "Labour"), "Other", as.character(winner_19)
      ),
       swing_angle=
       get_radians(map_scale(swing_con_lab,min_shift,max_shift,135,45)
      )
   ) |>
  ggplot()+
  geom_sf(aes(fill=elected), colour="#636363", alpha=.2, linewidth=.01)+
  geom_spoke(
    aes(x=bng_e, y=bng_n, angle=swing_angle, colour=elected, linewidth=is_flipped),
      radius=7000, position="center_spoke"
    )+
  coord_sf(crs=27700, datum=NA)+
  scale_linewidth_ordinal(range=c(.2,.5))+
  scale_colour_manual(values=party_colours)+
  scale_fill_manual(values=party_colours)
```

## Plot dot-density graphic

The session ends with some hidden code used to generate a dot-density map of 2019 GE result. As you will find throughout this book, the challenge is not in wrangling ggplot2 in order to draw dots on map. Rather it is the data generating procedures needed to create point locations sampled within the spatial areas of constituencies that correspond to votes.

In the code below for each element of the dot-density map, we generate point locations by random spatial sampling within each constituency. This may take some time, just under 1 minute, execute. If you are new to R and the tidyverse the code below looks quite formidable. I don't describe the code here, but we will be revisiting and reusing these sorts of code design patterns throughout the book.

```{r}
#| echo: true
#| eval: false
# Collect 2019 GE data from which dots are approximated.
vote_data <- bes_2019 |>   
  filter(ons_const_id!="S14000051") |> 
  mutate(other_vote_19=total_vote_19-(con_vote_19 + lab_vote_19)) |> 
  select(
    ons_const_id, constituency_name, region, con_vote_19, lab_vote_19, other_vote_19
    ) |> 
  pivot_longer(
    cols=con_vote_19:other_vote_19, names_to="party", values_to="votes"
    ) |> 
  mutate(
    party=str_extract(party, "[^_]+"),
    votes_dot=round(votes/1000,0)
  ) |>  
  filter(!is.na(votes_dot))

# This might take a bit of time to execute.
start_time <- Sys.time()
sampled_points <- 
  cons_outline |>   select(geometry, pcon21cd) |> filter(pcon21cd!="S14000051") |>  
  inner_join(
    vote_data |> 
    group_by(ons_const_id) |>  summarise(votes_dot=sum(votes_dot)) |>  ungroup()
    , by=c("pcon21cd"="ons_const_id")
    ) |> 
  nest(data=everything()) |> 
  mutate(
    sampled_points=
    map(
      data,
      ~sf::st_sample(
        x=.x,
        size=.x$votes_dot, exact=TRUE, type="random"
        ) |> 
        st_coordinates() |>   as_tibble(.name_repair=~c("east", "north"))
        ),
     const_id=map(data, ~.x |>  st_drop_geometry() |>   select(pcon21cd, votes_dot) |>   uncount(votes_dot))
    ) |> 
  unnest(-data) |> 
  select(-data)
end_time <- Sys.time()
end_time - start_time

point_votes <- vote_data |>   select(party, votes_dot) |>  uncount(votes_dot)
sampled_points  <- sampled_points |>  bind_cols(point_votes)
```

The ggplot2 code:

```{r}
party_colours <- c(con, lab, other)
sampled_points |> 
  ggplot() +
  geom_sf(data=cons_outline, fill="transparent", colour="#636363", linewidth=.03) +
  geom_sf(data=cons_outline |>  inner_join(vote_data, by=c("pcon21cd"="ons_const_id")) |> 
            group_by(region) |>  summarise(),
          fill="transparent", colour="#636363", linewidth=.1) +
  geom_point(aes(x=east,y=north, fill=party, colour=party), alpha=.5, size=.6, stroke=0)+
  scale_fill_manual(values=party_colours, "1 dot = 1,000 votes")+
  scale_colour_manual(values=party_colours, "1 dot = 1,000 votes")+
  guides(colour=guide_legend(override.aes=list(size=3)))+
  theme(
    plot.background = element_rect(fill="#ffffff", colour="#ffffff"), 
    legend.position ="right", axis.line = element_blank(), axis.text = element_blank(),
    axis.title.x = element_blank(), axis.title.y = element_blank()
    ) 
```
