---
title: "Chapter 04 Template"
author: "Roger Beecham"
format: html
---

## Introduction

Reopen your project and save this `.qmd` to the `reports` folder. When working through the materials from this chapter run each code chunk separately by clicking the small green arrow beside each code chunk. By default I have set `eval : false` on most code chunks, so when you knit the code chunk will not execute.

The technical element involves analysing [`stats19`](https://data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-safety-data) road crash data.

## Setup

The code block below points to the packages you need.

```{r setup}
#| echo: true
#| eval: false
library(tidyverse) # Bundle of packages for doing modern data analysis
library(sf) # Spatial operations
library(here) # For navigating project directory
library(fst) # For reading/writing large files
library(lubridate) # For working with dates
```

Bring in the pedestrian-vehicle crashses dataset (`ped_veh`) as an `.fst` file and supporting [IMD](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) and population datasets.

```{r import-sata}
#| echo: true
#| eval: false
url <- "https://vis4sds.github.io/data/ch4/ped_veh.fst"
download.file(url, here("data", "ped_veh.fst"))
ped_veh <- read_fst(here("data", "ped_veh.fst"))

url <- "https://vis4sds.github.io/data/ch4/imd.csv"
download.file(url, here("data", "imd.csv"))
imd <- read_csv(here("data", "imd.csv"))

url <- "https://vis4sds.github.io/data/ch4/lsoa_pop.csv"
download.file(url, here("data", "lsoa_pop.csv"))
lsoa_pop <- read_csv(here("data", "lsoa_pop.csv"))
```

Run the code block below to recode IMD to quintiles.

```{r recode-data}
#| echo: true
#| eval: false
# Recode imd into quintiles.
ped_veh <- ped_veh %>%
  mutate(
    driver_quintile=case_when(
      driver_imd_decile == "Most deprived 10%" ~ "1 most deprived",
      driver_imd_decile == "More deprived 10-20%" ~ "1 most deprived",
      driver_imd_decile == "More deprived 20-30%" ~ "2 more deprived",
      driver_imd_decile == "More deprived 30-40%" ~ "2 more deprived",
      driver_imd_decile == "More deprived 40-50%" ~ "3 mid deprived",
      driver_imd_decile == "Less deprived 40-50%" ~ "3 mid deprived",
      driver_imd_decile == "Less deprived 30-40%" ~ "4 less deprived",
      driver_imd_decile == "Less deprived 20-30%" ~ "4 less deprived",
      driver_imd_decile == "Less deprived 10-20%" ~ "5 least deprived",
      driver_imd_decile == "Least deprived 10%" ~ "5 least deprived",
      TRUE ~ driver_imd_decile),
    casualty_quintile=case_when(
      casualty_imd_decile == "Most deprived 10%" ~ "1 most deprived",
      casualty_imd_decile == "More deprived 10-20%" ~ "1 most deprived",
      casualty_imd_decile == "More deprived 20-30%" ~ "2 more deprived",
      casualty_imd_decile == "More deprived 30-40%" ~ "2 more deprived",
      casualty_imd_decile == "More deprived 40-50%" ~ "3 mid deprived",
      casualty_imd_decile == "Less deprived 40-50%" ~ "3 mid deprived",
      casualty_imd_decile == "Less deprived 30-40%" ~ "4 less deprived",
      casualty_imd_decile == "Less deprived 20-30%" ~ "4 less deprived",
      casualty_imd_decile == "Less deprived 10-20%" ~ "5 least deprived",
      casualty_imd_decile == "Least deprived 10%" ~ "5 least deprived",
      TRUE ~ casualty_imd_decile)
  )

# Recode IMD dataset, to be linked on crash location.
imd <- imd %>% mutate(
  quintile=ntile(decile,5),
  crash_quintile=case_when(
    quintile == 1 ~ "1 most deprived",
    quintile == 2 ~ "2 more deprived",
    quintile == 3 ~ "3 mid deprived",
    quintile == 4 ~ "4 less deprived",
    quintile == 5 ~ "5 least deprived"
  )
)

# Bring in IMD class of location. 
ped_veh <- ped_veh %>%
  # Crashes in Wales are excluded as English IMD
  inner_join(imd %>% select(lsoa_code, crash_quintile), by=c("lsoa_of_accident_location"="lsoa_code"))
```
