---
title: "Chapter 06 Template"
author: "Roger Beecham"
format: html
---

## Introduction

Reopen your project and save this `.qmd` file in the top-level folder. When working through the materials from this chapter run each code chunk separately by clicking the small green arrow beside each code chunk. By default I have set `eval: false` on most code chunks, so when you knit the code chunk will not execute.

The technical element involves analysing 2011 Census travel-to-work data.

## Setup

The code block below points to the packages you need. Uncomment the `intall_packages()` lines if you have not already downloaded `tidymodels`.

```{r}
#| echo: true
#| eval: false
library(tidyverse) 
library(sf) 
library(here) 
# Uncomment for download.
# install.packages("tidymodels") 
library(tidymodels) 
```

Bring in constituency-level data. 

```{r}
#| echo: true
#| eval: false
url <- "https://vis4sds.github.io/data/ch6/cons_data.csv"
download.file(url, here("data", "cons_data.csv"))
cons_data <- read_csv(here("data", "cons_data.csv"))

url <- "https://vis4sds.github.io/data/ch6/cons_hex.geojson"
download.file(url, here("data", "cons_hex.geojson"))
hex_data <- st_read(here("data", "cons_hex.geojson"))
```


## Plot parallel coordinates 

The data for the parallel coordinate plots is z-score transformed. In the code below we also identify majority Leave/Remain constituencies (`majority`), identify extreme Leave/Remain constituencies (`is_extreme`), select out the variables we need to plot and edit targeted variables in order to make the plot show Leave-related variables to the right. For example `degree`, a variable very negatively associated with leave, is given a negative sign. In order to encode variable types along an axis, we also need to `pivot_longer()`, so that each observation is a constituency, an explanatory variable and its corresponding tranformed variable value.  In the final `mutate()` the new `var` variable is given a factor ordering based on the largest to smallest associations with Leave -- see `ordered vars`.

```{r}
#| echo: true
#| eval: false

plot_data <- cons_data |> 
  mutate(
    across(c(leave, younger:heavy_industry), ~(.x-mean(.x))/sd(.x))
  ) |> 
  mutate(
    majority=if_else(leave>.5, "Leave", "Remain"),
    decile=ntile(leave, 10),
    is_extreme = decile > 9 | decile < 2
  )  |> 
  # Select out variables needed for plot.
  select(
    majority, is_extreme, constituency_name, leave, 
    degree, professional, younger, eu_born, no_car, white, own_home, christian, not_good_health, heavy_industry
    ) |>  
  # Change polarity in selected variables.
  mutate(degree=-degree, professional=-professional, younger=-younger, eu_born=-eu_born, no_car=-no_car) |>  
  # Gather explanatory variables for along rows.
  pivot_longer(cols= c(leave:not_good_health), names_to="var", values_to="z_score") |> 
  # Recode new explanatory variable as factor variable ordered according to known
  # assocs. Reverse order here as I've used coord_flip in plot.
  mutate(
    var=factor(var, levels=c("leave", order_vars)),
    var=fct_rev(var)
  ) 

order_vars <- 
  cons_data |>
  mutate(
    across(c(younger:heavy_industry), ~(.x-mean(.x))/sd(.x))
  ) |> 
  pivot_longer(cols=younger:heavy_industry, names_to="expl_var", values_to="prop") |> 
  group_by(expl_var) |>  
  summarise(cor=cor(leave,prop)) |>  ungroup() |>   arrange(cor) |>  
  pull(expl_var)

```

As ever the data processing code for the plot is more involved than the ggplot specification. We use `geom_path()` for drawing connected lines of the parallel coordinate plot, coloured according to `majority` and grouped by `constituency` -- that is, separate lines for each constituency.

```{r}
#| echo: true
#| eval: false
plot_data |>  
  ggplot()+
  geom_path(
    aes(x=var, y=z_score, group=c(constituency_name), colour=majority), 
    alpha=0.15, linewidth=.2
  )+
  scale_colour_manual(values=c("#b2182b","#2166ac")) +
  coord_flip()
```