---
title: "Chapter 07 Template"
author: "Roger Beecham"
format: html
---

## Introduction

Reopen your project and save this `.qmd` file in the top-level folder. When working through the materials from this chapter run each code chunk separately by clicking the small green arrow beside each code chunk. By default I have set `eval: false` on most code chunks, so when you knit the code chunk will not execute.


## Setup

The code block below points to the packages you need. Uncomment the `intall_packages()` lines if you have not already downloaded `ggdist`, `distributional` and `gganimate`.

```{r}
#| echo: true
#| eval: false
library(tidyverse) 
library(sf) 
library(here) 
library(tidymodels) 
library(rsample)

# install.packages("ggdist")
library(ggdist)
# install.packages("distributional")
library(distributional)
# install.packages("gganimate")
library(gganimate)
```

Bring in road crash data. If you have not previousyly download this, then uncomment the `download.file()`. Also, the `.geojson` file we use to draw cells of the risk theatre.

```{r}
#| echo: true
#| eval: false
# url <- "https://vis4sds.github.io/data/ch4/ped_veh.fst"
# download.file(url, here("data", "ped_veh.fst"))
ped_veh <- read_fst(here("data", "ped_veh.fst"))

url <- "https://vis4sds.github.io/data/ch7/risk_theatre.geojson"
download.file(url, here("data", "risk_theatre.geojson"))
```


